Parallel training by Pipeline Execution In AIRCoP

### 2021-12-12 report/paper.tex
proof read finished

%%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
%%\documentclass[sigplan,nonacm]{acmart}
\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
%%\setcopyright{acmcopyright}
\setcopyright{none}
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{9}
\acmDOI{}
%%\copyrightyear{2018}
%%\acmYear{2018}
%%\acmDOI{10.1145/1122445.1122456}

@@

\begin{abstract}
  Various approaches have been developed like data parallelism, model parallelism, attribute parallelism, and pipeline parallelism to train machine learning models in parallel, however, there are no existing frameworks that can integrate these approaches. Most commonly used machine learning frameworks such as TensorFlow and Pytorch only support a very small subset of parallel paradigms - the data parallelism and model parallelism. GPipe and PipeDream support different types of pipeline parallelism by dividing the model into several modules and running them in a pipeline fashion.
  AIRCop (Abstract IR for Comprehensive Parallelization) is a developing project that unifies all the parallelism paradigms mentioned above. Since all the machine learning models can be viewed as computation graphs, the parallelism paradigms (data parallelism, model parallelism, attribute parallelism, and pipeline parallelism) are just different kinds of computation graph transformation, ie transformation pass. AIRCop uses this idea and takes a machine learning model (basic sequential model) and parallelization annotations (describe the desired parallel behavior) as input and generates target program through different transformation passes.

@@

In the proposal, we mentioned that both stacked pipeline and queued pipeline will be supported by this project. However, we currently only support stacked pipeline for pipeline execution.\par
The reason is that we spent some time adding support for multiple-module for AIRCoP because AIRCoP originally only support single module. This involves changing the syntax of the frontend source language to make writing multiple-module models intuitive and easy. After finishing support for the stacked pipeline, We also found a severe problem in AIRCoP that will cause significant performance overhead - Memory Management. AIRCoP originally use cudaMalloc and cudaFree inside the training loop for memory allocation. Due to the complex synchronization mechanism of cudaMalloc/Free, these operations will cause a very long time. After analyzing the running time with NVIDIA profiler, we found that most of the execution time is spent on cudaMalloc/Free and it could be even 10x faster by pre-allocating the memory before the training loop. So we focus on addressing the memory management problem after finishing stacked pipeline. We currently use some code-motion techniques that will lift all the cudaMalloc out of the training loop for optimization.\par

@@

Although we only support stacked pipeline, it is sufficient for multi-module pipeline parallelism. We can still mix spatial partition with pipeline execution to explore different parallelism settings and gain some insight into parallel training as we will see in the evaluation.\par

@@

 Real-world machine learning applications are usually very large, they require a tremendous amount of data for training and can not work on a single device. As a result, various parallel strategies have been proposed for running large models in parallel among multiple distributed devices to achieve optimal performance.\par
 \subsection{Spatial Partition}There are different kinds of parallelism strategies, the most common one is Data parallelism\cite{krizhevsky2012imagenet}, where we split input across multiple devices and maintain the same network weight. Data parallelism is adapted and developed by most modern machine learning framework (TensorFlow\cite{abadi2016tensorflow}, Pytorch\cite{paszke2019pytorch}). Data parallelism requires network weight synchronizations between devices at the end of each iteration, therefore Data parallelism is ideal for a computation-intensive network with few weight parameters (convolution and pooling layer) but will introduce significant communication overhead for a network with a large number of weight parameters (fully-connected layer). Model parallelism\cite{dean2012large} is achieved by splitting the network weight tensor across devices while broadcasting the input tensor. Model parallelism reduces the communication overhead of Data Parallelism but is rarely used because there are often dependencies between network layers. Attribute parallelism\cite{jia2019beyond} can partition the attribute dimension in both input and weight parameters and require collective communication like AllReduce between operations. An experienced developer can combine different Parallelism strategies to achieve optimal performance. For example, \cite{krizhevsky2014one} uses data parallelism for the convolution layer and model parallelism for the fully-connected layer. All the above parallelism strategies can be categorized as Spatial Partition because they are based on the partition for data/weight/operation. \par

@@

 Different from Spatial Partition where input data/weight is distributed across multiple devices, there is another parallelism strategy: Pipeline Parallelism. Pipeline Parallelism divides a machine learning model into several modules with data dependency between them, divides a batch into several micro-batches, and runs them in parallel in a pipelined fashion. Both GPipe\cite{huang2019gpipe} and Pipedream\cite{narayanan2019pipedream} studied and developed Pipeline parallelism. Pipeline parallelism is especially useful when the training model is too big to fit inside a single device and there are dependencies between different layers. \par
 As shown in \cite{krizhevsky2014one}, combining different parallelism strategies can achieve better performance than only using one strategy. However, there is no unifying machine learning framework that can support all the parallelism strategies. TensorFlow\cite{abadi2016tensorflow} and Pytorch\cite{paszke2019pytorch} only supports Data Parallelism while GPipe\cite{huang2019gpipe} and Pipedream\cite{narayanan2019pipedream} supports Pipeline parallelism. Users are limited to only a subset of Parallelism strategies, and can not fully utilize the distributed training system. Different distributed system topologies also would require different strategies for optimal performance. As we mentioned before, Spatial Partition may introduce frequent tensor communication overhead between operations.

@@

\end{figure} This overhead may be negligible for a GPU Cluster with high-bandwidth interconnections but will cause a significant performance drop for large distributed systems consisting of several nodes of GPU clusters with low bandwidth connection between them\ref{fig:networktopo}. Because high bandwidth inter-GPU connection techniques like NV-Link can only scale to a limited number of devices, large training servers are often deployed across multiple nodes. It will be ideal to divide the model into several modules and assign each module into a cluster, use Spatial partition inside a GPU cluster and adopt pipeline parallelism for modules across different clusters.\par

@@

 AIRCoP \cite{aircopfeiwang} is the first framework that supports all the parallel parallelisms above thus allows users to fully explore different parallel settings. AIRCoP implemented spatial partition before and our work implemented Pipeline parallelism for AIRCoP, thus make AIRCoP the first unifying parallel training framework.\par

@@

 AIRCoP developed a formal syntax to represent tensor computation graphs. The syntax mainly has 3 components: Tensor and corresponding TensorType to represent input tensor and weight, Operation, and corresponding Operation Type to represent different Machine learning operations like Conv/MatMul/Relu and their input/output Type, Module to represent different modules. Operations and Modules are connected through let-bindings. And we use annotation of the formal syntax as an approach to formally represent different parallelism strategies. We developed annotation systems for both Tensor, Operation, and Modules. Tensor and Operation annotation represents Spatial Partition by indicating the dimension to partition. Module annotation indicates how many micro-batches the mini-batch is split into. Data parallelism can be achieved through annotating the input tensor indicating which dimension to split. Model parallelism can be achieved by annotating the weight tensor. Attribute parallelism can be achieved by a combination of tensor annotation and operation annotation. And pipeline parallelism can be simply done by annotation the Modules. Users only need to write the sequential forward pass of the Machine Learning model and the correct final training loop will be automatically computed through a series of transformations.

@@

 Pipeline Parallelism is ideal for distributed training workload across several nodes with weak inter-connection. Simply using Spatial Partition across different nodes will cause significant communication overhead between each operation. By splitting the training models into several modules and assigning each module to different nodes will limit the inter-node communication to the boundary of each module, therefore significantly reducing the inter-node communication overhead.\par

@@

 Simply splitting modules will cause large amount of stall time (or bubble time) due to complex dependencies between the sequential modules. As shown in\ref{fig:multiplemodule}, if we naively spilt the model into 4 modules $m_0,m_1,m_2,m_3$. There will be sequential dependencies between the forward pass $F_0, F_1, F_2, F_3$ of modules $m_0,m_1,m_2,m_3$, and reverse dependencies between backward pass $B_3, B_2, B_1, B_0$. The dependencies between the forward pass and backward pass will force strict ordering between modules, introduce synchronizations between module boundaries, and cause significant idle time on 4 devices. \par

@@

 Our approach is similar to GPipe\cite{huang2019gpipe}, we split a batch into multiple micro-batches and run the forward pass and the backward pass of multiple micro-batches sequentially. Only the forward pass and backward pass of the same micro-batches have dependencies between them. Different modules of different micro-batches could run in parallel as shown in\ref{fig:stackedpipeline}. Pipeline parallelism could allow users to fully utilize distributed devices by minimizing stall time caused by cross-module dependencies. \par
 Pipeline parallelism will insert blocking send/recv operations at the module boundaries to insure necessary execution ordering. And the gradient of multiple micro-batches will be accumulated after all the micro-batches finish and updated the weight synchronously for the entire mini-batch. Pipeline parallelism can easily scale to a large number of devices while maintaining ideal utilization of resources.\par
 Pipeline parallelism will introduce little communication overhead because it will only transfer data at the module boundaries rather than the frequent inter-operation tensor communication introduced by Spatial Partition. This feature of Pipeline parallelism makes it suitable not only for high-speed interconnection systems but also weakly connected systems. \par

@@

 Pipeline parallelism in AIRCoP is done through module annotation systems. The annotation consists of an integer representing the number of micro-batches (or pipelines). Users can split the model into several modules and annotate them for Pipeline parallelism. The resulting training loop is slightly different from naive multiple modules. Assume we have $n$ pipelines, the target program will have a top-level training loop for iterating mini-batches. Inside each mini-batch iteration, there will be a training loop of size $n$ for iterating the forward pass of the $n$ micro-batches and a training loop of size $n$ for iterating the backward pass of the $n$ micro-batches. The gradient will be accumulated inside each backward pass of micro-batches. After the backward pass training loop finishes, the weight will be updated based on gradient descent for the entire mini-batch.

@@

Transformation for pipeline parallelism has 2 stages; Collection Phase and Generation Phase. The Collection Phase collects the necessary information like forward/backward operation. The Generation Phase uses the collected information and the module annotation to generate the training loop.\par
During Collection Phase we need to collect the weight operations $E_w$, forward operations $E_f$ and backward operations $E_b$. For pipeline parallelism, we will add send/recv into $E_f$ and $E_b$ to maintain the data dependencies between modules. The corresponding variables of the operation is $xs^w$, $xs^f$, $xs^b$. We also need to calculate the lift variables $xs^l = Defined(E_f) \bigcap Used(E_b)$, the lifted variables $xs^l$ are the variables that are defined in the forward pass and used in the backward pass. Because we are generating different loops for forward pass and backward pass of micro-batches, we need to lift $xs^l$ into the outer scope and keep multiple versions of those variables for different micro-batches.\par
The formal description of the generation phase is as follows:

@@

  &G_f(\emptyset, xs^l)

@@

  &G_b(E^b, \emptyset)

@@

  op_l&([input],out\_type) \triangleq op(input^\epsilon, out\_type)

@@

$op_l$ means that the operation will read from the versioned lift variable instead of the originally variable. $xs^\epsilon = Zero^\epsilon$ will generates an array of $\epsilon$ versions of the variable $xs$.\par
$G_w$ generates the weight and their gradient and momentum, $G_l$ generate an array of lift nodes for different micro-batches, $G_f$ generates forward pass and write the correct value to the versioned lift variables, $G_b$ generates the backward pass and read from the versioned lift variable, $G_o$ generates the optimization pass. Because our source language does not allow mutation, we only need to write the versioned lift node at the end of the forward pass and the Generation phase can generate the correct training loop.

@@

 We implement AIRCoP and Pipeline Parallelism in LMS compiler framework\cite{rompf2010lightweight}, and generate CUDA\cite{sanders2010cuda} code with MPI\cite{gropp1999using}/NCCL for communication as the target executable program.\par
  LMS (Lightweight Modular Staging) is a library-based staging compiler that distinguishes stages solely based on Type information. LMS is extremely useful for developing DSL through overloading. It has a typed frontend making the code generation safe. LMS offers lifted (next-stage) versions of normal operators and normal control-flow statements make symbolic code generation easy and convenient. AIRCoP is simply a tensor computation DSL embedded inside LMS framework.\par

@@

 We use the transformation pass of LMS to transform the IR and simulate the tensor graph transformation to generate the correct training loop. We first need to perform DimName pass to merge the equivalent dimension of different tensors. For example, if matrix multiplication happens between matrixes of dimension $[m, n]$ and dimension $[j,k]$, then dimension $n$ and dimension $j$ can be merged.  Annotations of tensors and operations represent which dimension to partition, so we have to update those annotations as well. \par

@@

 The \textbf{Collection Phase} collects the forward operations, the weight nodes, and the corresponding backward operation of each node. The backward operation is collected in reverse order to maintain the correct semantic of differentiation. After Collection Phase, we have the necessary information needed to perform automatic differentiation (AD). To generate the correct training loop, we also need to collect lifted tensors. As we mentioned before, the forward pass and backward pass of micro-batches are iterated in different loops to achieve pipeline parallelism. Therefore, the tensors that are allocated in the forward pass and used in the backward pass will cause an error because forward and backward passes are now in different scopes. These tensors are called lifted tensors, we have to lift them into the outer scope and maintain multiple versions of them as an array of tensors. The forward and backward pass of micro-batches will use different versions of the lifted tensors. After collecting the lifted tensors, the Collection Phase is finished.\par
 The \textbf{Generation Phase} generates the correct training loop based on the collected information. It will add paired send/recv operations at the boundaries of different modules to maintain the correct data dependencies. Inside each module, it will first generate the top-level training loop for mini-batches. Then it will generate the forward training loop and backward training loop for micro-batches based on Module Annotations and use the collected backward operations to realize automatic differentiation. Inside both the forward and backward pass, the read and write to the lifted tensors will be transformed into read/write to the versioned tensor array. The weight gradient will be accumulated in the backward pass and optimized in the end for the entire mini-batch.\par
 The next pass will perform the Spatial Partition transformation. This part is of past work so we will not dig further into this pass. The main purpose of this pass is to partition the tensor and operations based on annotations and add the corresponding inter operation tensor communication for synchronization (AllReduce for MatrixMultiplication or Convolution).\par
 The final pass is to generate the corresponding MPI/NCCL instructions for inter-device communications. NCCL (NVIDIA Collective Communications Library) is the communication library introduced by NVIDIA for efficient collective and point-to-point communication across multiple GPU and multiple Nodes. Our target program is CUDA code with MPI/NCCL for communication, we use MPI to set up the environment and the NCCL communicator. We will create 2 NCCL communicators, one global communicator for cross-module communication (NCCLSend/Recv) and one local communicator for collective tensor communication (NCCLAllReduce) inside modules. We also need to handle tensor communication between modules if they have different partitions which can be naively implemented by gather and scatter. We use cuBLAS and cuDNN for matrix computation and convolution networks. To optimize memory management, we statically allocate GPU memory before the training loop and avoid time-exhausting cudaMalloc/cudaFree inside the training loop. \par
 We also use some code-motion techniques to lift the cudaMalloc inside the training loop and pre-allocate the memory before the training loop to optimize memory management. This could significantly improve performance because cudaMalloc/Free takes a very long time.\par

@@

 In this section, we show the evaluation benchmarks of various forms of Pipeline parallelism by running our generated CUDA code on 4 GeForce TITAN XP GPUs or server cuda.cs.purdue.edu. The server is a single GPU cluster with a high-speed inter-GPU connection. The ideal network topologies for Pipeline parallelism is a large training system consisting of several Nodes but due to limited hardware resources, we currently could not run the evaluation under that scenario. The purpose of this benchmark is to validate our graph transformation and gain a deeper insight into the choice of different parallelism strategies.\par
 We perform our test on a 12 convolution layer network for 40 iterations with different parallelism settings:

@@

   \item Mixed parallelism running on 4 GPUs, Naive 2 module parallelism with each module running in Data parallelism on 2 GPUs
   \item Mixed parallelism running on 4 GPUs, 2 modules in Pipelined parallelism (4 pipelines) with each module running in Data parallelism on 2 GPUs
   \item Mixed parallelism running on 4 GPUs, 2 modules in Pipelined parallelism (8 pipelines) with each module running in Data parallelism on 2 GPUs

@@

In conclusion, we implement Pipeline parallelism for AIRCoP, make it the first machine learning framework to integrate all the parallelism strategies, and offer user flexibility to fully explore different combinations of parallel training strategies under different hardware scenarios and gain a deeper insight into parallel training.

>>>>

%%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
%%\documentclass[sigplan,nonacm]{acmart}
\documentclass[sigplan, nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
%%\setcopyright{acmcopyright}
\setcopyright{none}
%%\acmJournal{PACMPL}
%%\acmVolume{1}
%%\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
%%\acmArticle{1}
%%\acmYear{2021}
%%\acmMonth{9}
%%\acmDOI{}
%%\copyrightyear{2018}
%%\acmYear{2018}
%%\acmDOI{10.1145/1122445.1122456}

@@

\begin{abstract}
  Various approaches have been developed like data parallelism, model parallelism, attribute parallelism, and pipeline parallelism to train machine learning models in parallel. However, there are no existing frameworks that can integrate these approaches. Most commonly used machine learning frameworks such as TensorFlow and Pytorch only support a tiny subset of parallel paradigms - the data parallelism and model parallelism. GPipe and PipeDream support different types of pipeline parallelism by dividing the model into several modules and running them in a pipeline fashion.
  AIRCop (Abstract IR for Comprehensive Parallelization) is a developing project that unifies all the parallelism paradigms mentioned above. Since all the machine learning models can be viewed as computation graphs, the parallelism paradigms (data parallelism, model parallelism, attribute parallelism, and pipeline parallelism) are just different kinds of computation graph transformation, i.e., transformation pass. AIRCop uses this idea and takes a machine learning model (basic sequential model) and parallelization annotations (describe the desired parallel behavior) as input and generates target program through different transformation passes.

@@

In the proposal, we mentioned that this project would support both stacked pipeline and queued pipeline. However, we currently only support stacked pipeline for pipeline execution.\par
The reason is that we spent some time adding support for multiple-module for AIRCoP because AIRCoP initially only supported single module. This work involves changing the syntax of the frontend to make writing multiple-module models intuitive and easy. After finishing support for the stacked pipeline, We also found a severe problem in AIRCoP that will cause significant performance overhead - Memory Management. AIRCoP initially uses cudaMalloc and cudaFree inside the training loop for memory allocation. Due to the complex synchronization mechanism of cudaMalloc/Free, these operations will cause a very long time. After analyzing the running time with NVIDIA profiler, we found that most of the execution time is spent on cudaMalloc/Free, and it could be even 10x faster by pre-allocating the memory before the training loop. So we focus on addressing the memory management problem after finishing stacked pipeline. We currently use some code-motion techniques that will lift all the cudaMalloc out of the training loop for optimization.\par

@@

Although we only support stacked pipeline, it is sufficient for multi-module pipeline parallelism. We can still mix spatial partition with pipeline execution to explore different parallelism settings and gain insight into parallel training, as we will see in the evaluation.\par

@@

 Real-world machine learning applications are usually tremendous. They require a tremendous amount of data for training and can not work on a single device. As a result, various parallel strategies have been proposed for running large models in parallel among multiple distributed devices to achieve optimal performance.\par
 \subsection{Spatial Partition}There are different kinds of parallelism strategies, the most common one is Data parallelism\cite{krizhevsky2012imagenet}, where we split input across multiple devices and maintain the same network weight. Data parallelism is adapted and developed by most modern machine learning framework (TensorFlow\cite{abadi2016tensorflow}, Pytorch\cite{paszke2019pytorch}). Data parallelism requires network weight synchronizations between devices at the end of each iteration. Therefore Data parallelism is ideal for a computation-intensive network with few weight parameters (convolution and pooling layer) but will introduce significant communication overhead for a network with a large number of weight parameters (fully-connected layer). Model parallelism\cite{dean2012large} is achieved by splitting the network weight tensor across devices while broadcasting the input tensor. Model parallelism reduces the communication overhead of Data Parallelism but is rarely used because there are often dependencies between network layers. Attribute parallelism\cite{jia2019beyond} can partition the attribute dimension in both input and weight parameters and require collective communication like AllReduce between operations. An experienced developer can combine different Parallelism strategies to achieve optimal performance. For example, \cite{krizhevsky2014one} uses data parallelism for the convolution layer and model parallelism for the fully-connected layer. The above parallelism strategies can be categorized as Spatial Partition because they are based on the partition for data/weight/operation. \par

@@

 Unlike Spatial Partition, where input data/weight is distributed across multiple devices, there is another parallelism strategy: Pipeline Parallelism. Pipeline Parallelism divides a machine learning model into several modules with data dependency between them, divides a batch into several micro-batches, and runs them in parallel in a pipelined fashion. Both GPipe\cite{huang2019gpipe} and Pipedream\cite{narayanan2019pipedream} studied and developed Pipeline parallelism. Pipeline parallelism is especially useful when the training system has low-bandwidth connections because it can limit the communication overhead to only happen at module boundaries.\par
 As shown in \cite{krizhevsky2014one}, combining different parallelism strategies can achieve better performance for different applications. However, no existing machine learning framework can support all the parallelism strategies. TensorFlow\cite{abadi2016tensorflow} and Pytorch\cite{paszke2019pytorch} only supports Data Parallelism while GPipe\cite{huang2019gpipe} and Pipedream\cite{narayanan2019pipedream} supports Pipeline parallelism. Users are limited to only a subset of Parallelism strategies and can not fully utilize the distributed training system.\par
  Different distributed system topologies also would require different strategies for optimal performance. As we mentioned before, Spatial Partition may introduce frequent tensor communication overhead between operations.

@@

\end{figure} This overhead may be negligible for a GPU Cluster with high-bandwidth interconnections but will cause a significant performance drop for large distributed systems consisting of several nodes of GPU clusters with low bandwidth connection between them Figure \ref{fig:networktopo}. Because high bandwidth inter-GPU connection techniques like NV-Link can only scale to a limited number of devices, large training servers are often deployed across multiple nodes. It will be ideal to divide the model into several modules, assign each module to a cluster, use Spatial partition inside a GPU cluster, and adopt pipeline parallelism for modules across different clusters.\par

@@

 AIRCoP \cite{aircopfeiwang} is the first framework that supports all the parallel parallelisms above, therefore allows users to fully explore different parallel settings. AIRCoP implemented spatial partition before and our work implemented Pipeline parallelism for AIRCoP, thus make AIRCoP the first unifying parallel training framework.\par

@@

 AIRCoP developed a formal syntax to represent tensor computation graphs. The syntax mainly has three components: Tensor and corresponding TensorType to represent input tensor and weight, Operation, and corresponding Operation Type to represent different Machine learning operations like Conv/MatMul/Relu and their input/output Type, Module to represent different modules. Operations and Modules are connected through let-bindings. Furthermore, we use annotation of the formal syntax to represent different parallelism strategies formally. We developed annotation systems for both Tensor, Operation, and Modules. Tensor and Operation annotation represents Spatial Partition by indicating the dimension to partition. Module annotation indicates how many micro-batches the mini-batch is split into. Users only need to write the sequential forward computation of the Machine Learning model, and the correct final training loop will be automatically computed through a series of transformations.

@@

 Pipeline Parallelism is ideal for distributed training workload across several nodes with weak inter-connection. Using Spatial Partition across different nodes will cause significant communication overhead between each operation. Splitting the training models into several modules and assigning each module to different nodes will limit the inter-node communication to the boundary of each module, therefore significantly reducing the inter-node communication overhead.\par

@@

 Simply splitting modules will cause large amount of stall time (or bubble time) due to complex dependencies between the sequential modules. As shown in Figure \ref{fig:multiplemodule}, if we naively spilt the model into 4 modules $m_0,m_1,m_2,m_3$. There will be sequential dependencies between the forward pass $F_0, F_1, F_2, F_3$ of modules $m_0,m_1,m_2,m_3$, and reverse dependencies between backward pass $B_3, B_2, B_1, B_0$. The dependencies between the forward pass and backward pass will force strict ordering between modules, introduce synchronizations between module boundaries, and cause significant idle time on 4 devices. \par

@@

\subsection{Stacked Pipeline Parallelism}
 Our approach is similar to GPipe\cite{huang2019gpipe}. We split a batch into multiple micro-batches and sequentially run the forward pass and the backward pass of multiple micro-batches. Because different micro-batches are stacked, we also call this Stacked Pipeline Parallelism. Only the forward and backward passes of the same micro-batches have dependencies between them. Different modules of different micro-batches could run in parallel as shown in Figure \ref{fig:stackedpipeline}. Stacked Pipeline parallelism allows users to fully utilize distributed devices by minimizing stall time caused by cross-module dependencies. \par
 Stacked Pipeline parallelism will insert blocking send/recv operations at the module boundaries to maintain necessary data dependencies. The gradient of multiple micro-batches will be accumulated and is used to update weight synchronously for the entire mini-batch. Stacked Pipeline parallelism can easily scale to a large number of devices while maintaining ideal utilization of resources.\par
 Stacked Pipeline parallelism will introduce little communication overhead because it will only transfer data at the module boundaries rather than the frequent inter-operation tensor communication introduced by Spatial Partition. This feature of Stacked Pipeline parallelism makes it suitable for both high-speed interconnection systems and weakly connected systems. \par
 Throughout the rest of this paper, the term Pipeline parallelism refers to Stacked Pipeline parallelism.
 \subsection{Pipeline Parallelism in AIRCoP}
 Pipeline parallelism in AIRCoP is done through module annotation systems. The annotation consists of an integer representing the number of micro-batches (or pipelines).
 Users can split the model into several modules and annotate them for Pipeline parallelism. The resulting training loop is different from naive multiple modules.

@@

Assume we have $\epsilon$ pipelines. The target program will have a top-level training loop for iterating mini-batches. Inside each mini-batch iteration, there will be a training loop of size $\epsilon$ for iterating the forward passes of the $\epsilon$ micro-batches and a training loop of size $\epsilon$ for iterating the backward passes of the $\epsilon$ micro-batches. The gradient will be accumulated inside each backward pass of micro-batches. After the backward pass training loop finishes, the weight will be updated based on gradient descent for the entire mini-batch.

@@

Transformation for pipeline parallelism has 2 stages; \textbf{Collection Phase} and \textbf{Generation Phase}. The Collection Phase collects the necessary information like forward/backward operation. The Generation Phase uses the collected information and the module annotation to generate the training loop.\par
\paragraph{Collection Phase}
During Collection Phase we need to collect the weight operations $E_w$, forward operations $E_f$ and backward operations $E_b$. For pipeline parallelism, we will add send/recv into $E_f$ and $E_b$ to maintain the data dependencies between modules. The corresponding variables of the operation is $xs^w$, $xs^f$, $xs^b$. We also need to calculate the lift variables $xs^l = Defined(E_f) \bigcap Used(E_b)$, the lifted variables $xs^l$ are the variables defined in the forward pass and used in the backward pass. Because we are generating different loops for forward passes and backward passes of micro-batches, we need to lift $xs^l$ into the outer scope and keep multiple versions of those variables for different micro-batches.\par
\paragraph{Generation Phase}
The formal description of the Generation Phase is as follows:

@@

  &G_f(\emptyset, xs^l)\\

@@

  &G_b(E^b, \emptyset)\\

@@

  \\
  op_l&([input],out\_type) \triangleq op([input^\epsilon], out\_type)

@@

$op_l$ means that the operation will read from the versioned lift variable instead of the original variable. $xs^\epsilon = Zero^\epsilon$ will generates an array of $\epsilon$ versions of the variable $xs$.\par
$G_w$ generates the weight and their gradient and momentum, $G_l$ generate an array of lift nodes for different micro-batches, $G_f$ generates forward pass and write the correct value to the versioned lift variables, $G_b$ generates the backward pass and read from the versioned lift variable, $G_o$ generates the optimization pass. Because our source language does not allow mutation, all the mutation operations (Assign/Accumulate/Optimize) are automatically generated in the Generation Phase. We only need to write the versioned lift node at the end of the forward pass, and the Generation phase can generate the correct training loop.

@@

We implement AIRCoP and Stacked Pipeline Parallelism in LMS compiler framework\cite{rompf2010lightweight}, and generate CUDA\cite{sanders2010cuda} code with MPI\cite{gropp1999using}/NCCL for communication as the target executable program.\par
\subsection{Integration into LMS}
 LMS (Lightweight Modular Staging) is a library-based staging compiler that distinguishes stages solely based on Type information. LMS is beneficial for developing DSL through overloading. It has a typed frontend making the code generation safe. LMS offers lifted (next-stage) versions of regular operators, and standard control-flow statements make symbolic code generation easy and convenient. AIRCoP is simply a tensor computation DSL embedded inside LMS framework.\par

@@

\subsection{Training loop transformation For Pipeline Parallelism}
We use the transformation pass of LMS to transform the IR and simulate the tensor graph transformation to generate the correct training loop. We first need to perform DimName pass to merge the equivalent dimension of different tensors. For example, if matrix multiplication happens between matrixes of dimension $[m, n]$ and dimension $[j,k]$, then dimension $n$ and dimension $j$ can be merged.  Annotations of tensors and operations represent which dimension to partition, so we have to update those annotations too. \par

@@

\paragraph{Collection Phase}
The Collection Phase collects the forward operations, the weight nodes, and the corresponding backward operation of each node. The backward operation is collected in reverse order to maintain the correct semantic of differentiation. Because the backward pass should be generated after the forward pass, we collect backward operation in the form of closures. We implement the Collection Phase by defining a function aircopCollect for each IR Node to collect the corresponding information. After the Collection Phase, we have the necessary information to perform automatic differentiation (AD). We also need to collect lifted tensors to generate the correct training loop. As we mentioned before, the forward pass and backward pass of micro-batches are iterated in different loops to achieve pipeline parallelism. Therefore, the tensors allocated in the forward pass and used in the backward pass will cause an error because forward and backward passes are now in different scopes. These tensors are called lifted tensors. We have to lift them into the outer scope and maintain multiple versions of them as an array of tensors. Micro-batches' forward and backward passes will use different versions of the lifted tensors. After collecting the lifted tensors, the Collection Phase is finished.\par
\paragraph{Generation Phase}
The Generation Phase generates the correct training loop based on the collected information. It will add paired send/recv operations at the boundaries of different modules to maintain the correct data dependencies. Inside each module, it will first generate the top-level training loop for mini-batches. Then it will generate the forward training loop and backward training loop for micro-batches based on Module Annotations and use the collected backward operations to realize automatic differentiation as we mentioned in Section \ref{formalpipe}. Inside both the forward and backward pass, the read and write to the lifted tensors will be transformed into read/write to the versioned tensor array. The weight gradient will be accumulated in the backward pass and optimized at the end of the entire mini-batch.\par
\subsection{Spatial Partition Pass}
The next pass will perform the Spatial Partition transformation. This part is of past work, so we will not dig further into this pass. The main purpose of this pass is to partition the tensor and operations based on annotations and add the corresponding inter operation tensor communication for synchronization (AllReduce for MatrixMultiplication or Convolution).\par
\subsection{MPI/NCCL Pass}
The final pass is to generate the corresponding MPI/NCCL instructions for inter-device communications. NCCL (NVIDIA Collective Communications Library) is the communication library introduced by NVIDIA for efficient collective and point-to-point communication across multiple GPU and multiple Nodes. Our target program is CUDA code with MPI/NCCL for communication. We use MPI to set up the environment and the NCCL communicator. We will create 2 NCCL communicators, one global communicator for cross-module communication (NCCLSend/Recv) and one local communicator for collective tensor communication (NCCLAllReduce) inside modules. We also need to handle tensor communication between modules if they have different Spatial partitions, which can be naively implemented by gather and scatter. We use cuBLAS and cuDNN for matrix computation and convolution networks.\par
We statically allocate GPU memory before the training loop and avoid time-exhausting cudaMalloc/cudaFree inside the training loop to optimize memory management. We currently use code-motion techniques to lift the cudaMalloc inside the training loop and pre-allocate the memory before the training loop to optimize memory management. This approach could significantly improve performance because cudaMalloc/Free often takes a very long time.\par

@@

 This section shows the evaluation benchmarks of various forms of Pipeline parallelism by running our generated CUDA code on 4 GeForce TITAN XP GPUs or server cuda.cs.purdue.edu. The server is a single GPU cluster with a high-speed inter-GPU connection. The ideal network topology for Pipeline parallelism is an extensive training system consisting of several Nodes, but we currently could not run the evaluation under that scenario due to limited hardware resources and time. The purpose of this benchmark is to validate our graph transformation and gain a deeper insight into the choice of different parallelism strategies.\par
 We perform our test on a 12 layer convolution network for 40 iterations with different parallelism settings:

@@

   \item Mixed parallelism running on 4 GPUs --- Naively partitioned into 2 modules with each module running in Data parallelism on 2 GPUs
   \item Mixed parallelism running on 4 GPUs --- 2 modules in Stacked Pipeline parallelism (4 pipelines) with each module running in Data parallelism on 2 GPUs
   \item Mixed parallelism running on 4 GPUs --- 2 modules in Stacked Pipeline parallelism (8 pipelines) with each module running in Data parallelism on 2 GPUs

@@

In conclusion, we implement Pipeline parallelism for AIRCoP, make it a machine learning framework that can integrate different parallelism strategies, and offer user flexibility to fully explore different combinations of parallel training strategies under different hardware scenarios and gain a deeper insight into parallel training.
